---
title: "STAT 1341 - Sports Analytics - Midterm Project"
author: "Zach Palmer"
subtitle: 'Due: 11/7/21'
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## Environment Setup and Packages

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(stats)
library(ggplot2)
library(gridExtra)
library(dplyr)
```

## Data Collection 

### Position Players
Read in the csv file of all qualified (according to FanGraphs) batters from the '06 to '21 seasons and manipulate/clean accordingly to reach the desired format.

```{r}
batting_stats = read.csv("data/batting_statistics_06_21.csv")
batting_stats = batting_stats %>%
  arrange(Name, Age)
```

```{r}
# we only want to keep entries that have the same players stats in back to back years
keep = NULL

# iterate through the data frame and check if the previous entry is the same player 
# but with his stats from the prior season (one year younger), if so designate the row to be kept
for (i in 2:nrow(batting_stats)) {
  if (batting_stats[i, "Name"] == batting_stats[i - 1, "Name"] && 
      batting_stats[i, "Age"] == (batting_stats[i - 1, "Age"] + 1)) {
    keep = c(keep, i)
  }
}

# iterate through the data frame and check if the next entry is the same player 
# but with his stats from the next season (one year older), if so designate the row to be kept
for (i in 1:(nrow(batting_stats) - 1)) {
  if (batting_stats[i, "Name"] == batting_stats[i + 1, "Name"] && 
      batting_stats[i, "Age"] == (batting_stats[i + 1, "Age"] - 1)) {
    keep = c(keep, i)
  }
}

# remove duplicates and sort
keep = unique(keep)
keep = sort(keep)

# only keep the designated rows, remove any rows with missing values, 
# and select the columns we are interested in
batting_subset = batting_stats[keep,] %>%
  drop_na() %>%
  select(ï..Season, Name, Age, wOBA, Bat, BsR, Def, WAR)
```

### Pitchers\   
Read in the csv file of all qualified (according to FanGraphs) pitchers from the '06 to '21 seasons and manipulate/clean accordingly to reach the desired format.

```{r}
pitching_stats = read.csv("data/pitching_statistics_06_21.csv")
pitching_stats = pitching_stats %>%
  arrange(Name, Age)
```

```{r}
# we only want to keep entries that have the same players' stats in back to back years
keep = NULL

# iterate through the data frame and check if the previous entry is the same player 
# but with his stats from the prior season (one year younger), if so designate the row to be kept
for (i in 2:nrow(pitching_stats)) {
  if (pitching_stats[i, "Name"] == pitching_stats[i - 1, "Name"] && 
      pitching_stats[i, "Age"] == (pitching_stats[i - 1, "Age"] + 1)) {
    keep = c(keep, i)
  }
}

# iterate through the data frame and check if the next entry is the same player 
# but with his stats from the next season (one year older), if so designate the row to be kept
for (i in 1:(nrow(pitching_stats) - 1)) {
  if (pitching_stats[i, "Name"] == pitching_stats[i + 1, "Name"] && 
      pitching_stats[i, "Age"] == (pitching_stats[i + 1, "Age"] - 1)) {
    keep = c(keep, i)
  }
}

# remove duplicates and sort
keep = unique(keep)
keep = sort(keep)

# only keep the designated rows, remove any rows with missing values, 
# and select the columns we are interested in
pitching_subset = pitching_stats[keep,] %>%
  drop_na() %>%
  select(ï..Season, Name, Age, vFA..pi., WHIP, xFIP, WAR)
```

## Descriptive Statistics

### Position Players\  
Create both a table of summary statistics and visualizations for the descriptive statistics for batters.

```{r}
# create a single row data frame that summarizes the batting statistics 
# the data frame is a list of the mean, median, and sd of each of the individual stats 
batting_summary = batting_subset %>%
  summarise(
    meanAge = mean(Age),
    medianAge = median(Age),
    sdAge = sd(Age),
    meanWOBA = mean(wOBA),
    medianWOBA = median(wOBA),
    sdWOBA = sd(wOBA),
    meanBat = mean(Bat),
    medianBat = median(Bat),
    sdBat = sd(Bat),
    meanBsR = mean(BsR),
    medianBsR = median(BsR),
    sdBsR = sd(BsR),
    meanDef = mean(Def),
    medianDef = median(Def),
    sdDef = sd(Def),
    meanWAR = mean(WAR),
    medianWAR = median(WAR),
    sdWAR = sd(WAR)
  )
batting_summary
```

```{r fig.height=12, fig.width=9}
# create a scatterplot of Age vs each statistic
# create a histogram showing the distribution of each of the statistics
par(mfrow = c(5, 2))

#wOBA
plot(batting_subset$Age, batting_subset$wOBA, xlab = "Age", ylab = "wOBA")
hist(batting_subset$wOBA, main = "Distribution of wOBA", xlab = "wOBA")

#Bat
plot(batting_subset$Age, batting_subset$Bat, xlab = "Age", ylab = "Bat")
hist(batting_subset$Bat, main = "Distribution of Batting Runs", xlab = "Bat")

#BsR
plot(batting_subset$Age, batting_subset$BsR, xlab = "Age", ylab = "BsR")
hist(batting_subset$BsR, main = "Distribution of Baserunning Runs", xlab = "BsR")

#Def
plot(batting_subset$Age, batting_subset$Def, xlab = "Age", ylab = "Def")
hist(batting_subset$Def, main = "Distribution of Defensive Runs", xlab = "Def")

#WAR
plot(batting_subset$Age, batting_subset$WAR, xlab = "Age", ylab = "WAR")
hist(batting_subset$WAR, main = "Distribution of WAR", xlab = "WAR")
```

### Pitchers\  
Create both a table of summary statistics and visualizations for the descriptive statistics for pitchers.

```{r}
# create a single row data frame that summarizes the pitching statistics 
# the data frame is a list of the mean, median, and sd of each of the individual stats 
pitching_summary = pitching_subset %>%
  summarise(
    meanAge = mean(Age),
    medianAge = median(Age),
    sdAge = sd(Age),
    meanAvgVelo = mean(vFA..pi.),
    medianAvgVelo = median(vFA..pi.),
    sdAvgVelo = sd(vFA..pi.),
    meanWHIP = mean(WHIP),
    medianWHIP = median(WHIP),
    sdWHIP = sd(WHIP),
    meanXFIP = mean(xFIP),
    medianXFIP = median(xFIP),
    sdXFIP = sd(xFIP),
    meanWAR = mean(WAR),
    medianWAR = median(WAR),
    sdWAR = sd(WAR)
  )
pitching_summary
```

```{r fig.height=12, fig.width=9}
# create a scatterplot of Age vs each statistic
# create a histogram showing the distribution of each of the statistics
par(mfrow = c(4, 2))

#Avg 4-Seam Velo
plot(pitching_subset$Age, pitching_subset$vFA..pi., xlab = "Age", ylab = "vFA..pi.")
hist(pitching_subset$vFA..pi., main = "Distribution of Avg. 4-Seam Velocity", xlab = "vFA..pi.")

#WHIP
plot(pitching_subset$Age, pitching_subset$WHIP, xlab = "Age", ylab = "WHIP")
hist(pitching_subset$WHIP, main = "Distribution of WHIP", xlab = "WHIP")

#xFIP
plot(pitching_subset$Age, pitching_subset$xFIP, xlab = "Age", ylab = "xFIP")
hist(pitching_subset$xFIP, main = "Distribution of xFIP", xlab = "xFIP")

#WAR
plot(pitching_subset$Age, pitching_subset$WAR, xlab = "Age", ylab = "WAR")
hist(pitching_subset$WAR, main = "Distribution of WAR", xlab = "WAR")
```

## Inferential Statistics  

### Position Players\  
Generate the aging curves for the position players

```{r}
# create an empty data frame with 6 columns to fill with the age ranges and 
# the differences in the statistic values between age changes/years 
aging_diffs = data.frame(matrix(ncol = 6, nrow = 0))
cols = c("ageRange", "wOBA_diff", "Bat_diff", "BsR_diff", "Def_diff", "WAR_diff")
colnames(aging_diffs) = cols
```

```{r}
# iterate over the data frame and for each set of consecutive seasons calculate the
# difference between the statistic values for the following year and the current
for (i in 1:(nrow(batting_subset) - 1)) {
  # check that the next row represents the following season for the same player
  if (batting_subset[i, "Name"] == batting_subset[i + 1, "Name"] && 
      batting_subset[i, "Age"] == (batting_subset[i + 1, "Age"] - 1)) {
    # grab each of the values
    ageRange = paste(batting_subset[i, "Age"], batting_subset[i + 1, "Age"], sep = "-")
    wOBA_diff = batting_subset[i + 1, "wOBA"] - batting_subset[i, "wOBA"]
    Bat_diff = batting_subset[i + 1, "Bat"] - batting_subset[i, "Bat"]
    BsR_diff = batting_subset[i + 1, "BsR"] - batting_subset[i, "BsR"]
    Def_diff = batting_subset[i + 1, "Def"] - batting_subset[i, "Def"]
    WAR_diff = batting_subset[i + 1, "WAR"] - batting_subset[i, "WAR"]
    # put the values together into a list representing a row in the data frame
    row = list(ageRange, wOBA_diff, Bat_diff, BsR_diff, Def_diff, WAR_diff)
    # add the row to the data frame
    aging_diffs[nrow(aging_diffs) + 1,] = row
  }
}
```

```{r}
# create a new data frame that contains the average difference between each 
# of the statistic values for each age range
sum_of_diffs = aging_diffs %>%
  # group the data frame by the age ranges so summarise() takes the mean for each
  group_by(ageRange) %>%
  summarise(
    N = n(),
    avg_wOBA_diff = mean(wOBA_diff),
    avg_Bat_diff = mean(Bat_diff), 
    avg_BsR_diff = mean(BsR_diff), 
    avg_Def_diff = mean(Def_diff), 
    avg_WAR_diff = mean(WAR_diff)
  )
```

```{r}
# outputs the sample size for each of the age ranges, just useful information to know
print(tibble(sum_of_diffs[, 1:2]), n = 21)
```

```{r fig.height=12, fig.width=9}
# plot the aging curves for each of the statistics
par(mfrow = c(1, 2))

WAR = ggplot(aes(x = ageRange, y = avg_WAR_diff, group = 1), data = sum_of_diffs) + 
  labs(title = "Aging Curve for Position Player WAR", 
       x = "Age Range", y = "Change in WAR") +
  geom_point() + 
  geom_line(color = "blue") +
  scale_y_continuous(breaks = seq(-2.0, 1.0, 0.2)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

wOBA = ggplot(aes(x = ageRange, y = avg_wOBA_diff, group = 1), data = sum_of_diffs) + 
  labs(title = "Aging Curve for wOBA", 
       x = "Age Range", y = "Change in wOBA") +
  geom_point() + 
  geom_line(color = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

Bat = ggplot(aes(x = ageRange, y = avg_Bat_diff, group = 1), data = sum_of_diffs) + 
  labs(title = "Aging Curve for Batting Runs", 
       x = "Age Range", y = "Change in Batting Runs") +
  geom_point() + 
  geom_line(color = "blue")  +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

BsR = ggplot(aes(x = ageRange, y = avg_BsR_diff, group = 1), data = sum_of_diffs) + 
  labs(title = "Aging Curve for Baserunning Runs", 
       x = "Age Range", y = "Change in Baserunning Runs") +
  geom_point() + 
  geom_line(color = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

Def = ggplot(aes(x = ageRange, y = avg_Def_diff, group = 1), data = sum_of_diffs) + 
  labs(title = "Aging Curve for Defensive Runs", 
       x = "Age Range", y = "Change in Defensive Runs") +
  geom_point() + 
  geom_line(color = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

grid.arrange(wOBA, Bat, BsR, Def, WAR, ncol = 2)
```

### Pitchers\  
Generate the aging curves for the pitchers

```{r}
# create an empty data frame with 5 columns to fill with the age ranges and 
# the differences in the statistic values between age changes/years 
pitching_aging_diffs = data.frame(matrix(ncol = 5, nrow = 0))
cols = c("ageRange", "vFA_diff", "WHIP_diff", "xFIP_diff", "WAR_diff")
colnames(pitching_aging_diffs) = cols
```

```{r}
# iterate over the data frame and for each set of consecutive seasons calculate the
# difference between the statistic values for the following year and the current
for (i in 1:(nrow(pitching_subset) - 1)) {
  # check that the next row represents the following season for the same player
  if (pitching_subset[i, "Name"] == pitching_subset[i + 1, "Name"] && 
      pitching_subset[i, "Age"] == (pitching_subset[i + 1, "Age"] - 1)) {
    # grab each of the values
    ageRange = paste(pitching_subset[i, "Age"], pitching_subset[i + 1, "Age"], sep = "-")
    vFA_diff = pitching_subset[i + 1, "vFA..pi."] - pitching_subset[i, "vFA..pi."]
    WHIP_diff = pitching_subset[i + 1, "WHIP"] - pitching_subset[i, "WHIP"]
    xFIP_diff = pitching_subset[i + 1, "xFIP"] - pitching_subset[i, "xFIP"]
    WAR_diff = pitching_subset[i + 1, "WAR"] - pitching_subset[i, "WAR"]
    # put the values together into a list representing a row in the data frame
    row = list(ageRange, vFA_diff, WHIP_diff, xFIP_diff, WAR_diff)
    # add the row to the data frame
    pitching_aging_diffs[nrow(pitching_aging_diffs) + 1,] = row
  }
}
```

```{r}
# create a new data frame that contains the average difference between each 
# of the statistic values for each age range
pitching_sum_of_diffs = pitching_aging_diffs %>%
  # group the data frame by the age ranges so summarise() takes the mean for each
  group_by(ageRange) %>%
  summarise(
    N = n(),
    avg_vFA_diff = mean(vFA_diff),
    avg_WHIP_diff = mean(WHIP_diff), 
    avg_xFIP_diff = mean(xFIP_diff), 
    avg_WAR_diff = mean(WAR_diff)
  )
```

```{r}
# outputs the sample size for each of the age ranges, just useful information to know
print(tibble(pitching_sum_of_diffs[, 1:2]), n = 25)
```

```{r fig.height=8, fig.width=9}
# plot the aging curves for each of the statistics
par(mfrow = c(1, 2))

pitch_WAR = ggplot(aes(x = ageRange, y = avg_WAR_diff, group = 1), data = pitching_sum_of_diffs) + 
  labs(title = "Aging Curve for Pitcher WAR", 
       x = "Age Range", y = "Change in WAR") +
  geom_point() + 
  geom_line(color = "blue") +
  scale_y_continuous(breaks = seq(-2.0, 1.0, 0.2)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

vFA = ggplot(aes(x = ageRange, y = avg_vFA_diff, group = 1), data = pitching_sum_of_diffs) + 
  labs(title = "Aging Curve for Average 4-Seam Velocity", 
       x = "Age Range", y = "Change in Avg. 4-Seam Velo") +
  geom_point() + 
  geom_line(color = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

WHIP = ggplot(aes(x = ageRange, y = avg_WHIP_diff, group = 1), data = pitching_sum_of_diffs) + 
  labs(title = "Aging Curve for WHIP", 
       x = "Age Range", y = "Change in WHIP") +
  geom_point() + 
  geom_line(color = "blue")  +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

xFIP = ggplot(aes(x = ageRange, y = avg_xFIP_diff, group = 1), data = pitching_sum_of_diffs) + 
  labs(title = "Aging Curve for xFIP", 
       x = "Age Range", y = "Change in xFIP") +
  geom_point() + 
  geom_line(color = "blue") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

grid.arrange(vFA, WHIP, xFIP, pitch_WAR, ncol = 2)
```